---
title: "The IndexedFst class"
author: Pierre-Luc Germain
package: scanMiR
output:
  BiocStyle::html_document
abstract: |
  Indexed wrapper for fst objects, enabling indexed fast random access to on-disk data.frames or GRanges
vignette: |
  %\VignetteIndexEntry{4_IndexedFst}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(BiocStyle)
library(scanMiR)
```

# IndexedFst

The `IndexedFst` class provides fast named random access to indexed `fst` files. It is based on the `r CRANpkg("fst")` package, which provides fast random reading of data frames. This is particularly useful to manipulate large collections of binding sites without loading them all in memory.

Creating an indexed fst file from a data.frame is very simple:

```{r}
# we create a temporary directory in which the file will be saved
tmp <- tempdir()
f <- file.path(tmp, "test")
# we create a dummy data.frame
d <- data.frame( category=sample(LETTERS[1:4], 10000, replace=TRUE),
                 var2=sample(LETTERS, 10000, replace=TRUE),
                 var3=runif(10000) )

saveIndexedFst(d, index.by="category", file.prefix=f)
```

The file can then be loaded (without having all the data in memory) in the following way:

```{r}
d2 <- loadIndexedFst(f)
class(d2)
summary(d2)
```

We can see that `d2` is considerably smaller than the original `d`:

```{r}
format(object.size(d),units="Kb")
format(object.size(d2),units="Kb")
```

Nevertheless, a number of functions can be used normally on the object:

```{r}
nrow(d2)
ncol(d2)
colnames(d2)
head(d2)
```

In addition, the object can be accessed as a list (using the indexed variable). Since in this case the file is indexed using the category column, the different categories can be accessed as `names` of the object:

```{r}
names(d2)
lengths(d2)
```
We can read specifically the rows pertaining to one category using:

```{r}
catB <- d2$B
head(catB)
```

# Storing GRanges as IndexedFst

In addition to data.frames, [GRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) can be saved as indexed Fst. To demonstrate this, we first create a dummy GRanges object:

```{r}
gr <- GRanges(sample(LETTERS[1:3],200,replace=TRUE), IRanges(seq_len(200), width=2))
gr$propertyA <- factor(sample(letters[1:5],200,replace=TRUE))
gr
```
Again the file can then be loaded (without having all the data in memory) in the following way:

```{r}
f2 <- file.path(tmp, "test2")
saveIndexedFst(gr, index.by="seqnames", file.prefix=f2)
d1 <- loadIndexedFst(f2)
names(d1)
head(d1$A)
```
Similarly, we could index using a different column:

```{r}
saveIndexedFst(gr, index.by="propertyA", file.prefix=f2)
d2 <- loadIndexedFst(f2)
names(d2)
```

# More...

## Multithreading

The `r CRANpkg("fst")` package supports multithreaded reading and writing. This can also be applied for `IndexedFst`, using the `nthreads` argument of `loadIndexedFst` and `saveIndexedFst`.

## Under the hood

The `IndexedFst` class is simply a wrapper around the `fst` package. In addition to the `fst` file, an `rds` file is saved containing the index data. For example, for our last example, the following files have been saved:

```{r}
list.files(tmp, "test2")
```
Either file (or the prefix) can be used for loading, but both files need to have the same prefix.


<br/><br/>

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
